---
title: "hawksDA"
author: "Lawrence Jang"
date: "5/26/2023"
output: html_document
---

```{r}
library(tidyverse)
library(ggplot2)
```

## Data

Read in the Kaggle dataset.

```{r}
# read in dataset
df = read.csv("nbaplayersdraft.csv")
```


All the variables for the dataset.

```{r}
# get variable names
colnames(df)
```

First couple rows for the dataset. 

```{r}
# display first few rows
head(df)
```

## Part 1: Data Comprehension

(A) Which NBA team(s) has drafted the most players who...
a. went to Duke and were drafted in or before the 2000 draft?
b. have a first name that begins with D and were drafted in an even year draft (1990, 1992, 1994, ...)?

```{r}
# part A
# filter for drafts before 2001 and players from Duke
a = df %>% filter(year <= 2000 & college == "Duke") 
# display values by frequency
sort(table(a$team), decreasing = TRUE)
```

Dallas, Minnesota, and Phoenix drafted the most players who went to Duke and 
were drafted in or before the 2000 draft.

```{r}
# part B
# filter for names that start with D and even drafts
dEven = df %>% filter(grepl("^D", player) & year %% 2 == 0)
# display values by frequency
sort(table(dEven$team), decreasing = TRUE)
```

Boston, Milwaukee, and Seattle drafted the most players who went to Duke and 
were drafted in or before the 2000 draft.


(B) Describe the relationship between a team’s first round pick slot in one year with their first-round pick slot in
the subsequent year.

```{r}
# get all first round picks, adjusting the first round pick limit based on expansion/number of teams in league
firstRoundPicks =  df %>% filter((overall_pick <= 30 & year >= 2004)|(overall_pick <= 29 & year >= 1995 & year < 2004)|(overall_pick <= 27 & year >= 1989 & year < 1995)) 
# keep the highest pick for a team in a given year
singleFRPs = firstRoundPicks %>% 
  group_by(year, team) %>%
  slice_min(overall_pick) %>%
  ungroup()
# get pick and next year pick pairs
pickComps <- singleFRPs %>%
  group_by(team) %>%
  arrange(year) %>%
  mutate(next_year_pick = lead(overall_pick)) %>%
  filter(!is.na(next_year_pick)) %>%
  select(year, team, overall_pick, next_year_pick)
```

Summary statistics on difference between a team’s first round pick slot in one year with their first-round pick slot in the subsequent year.

```{r}
pickComps$diff = pickComps$next_year_pick - pickComps$overall_pick
# correlation
c = cor( pickComps$next_year_pick, pickComps$overall_pick)
# mean
mea = mean(pickComps$diff)
# median
med = median(pickComps$diff)
# standard deviation
std = sd(pickComps$diff)
print(paste("Correlation:", c))
print(paste("Mean:", mea))
print(paste("Median:", med))
print(paste("Standard Deviation:", std))
```                                                                                                                    

```{r}
ggplot(pickComps, aes(x = overall_pick, y = next_year_pick)) +
  geom_point() +
  labs(x = "First Round Pick Slot (Current Year)",
       y = "First Round Pick Slot (Next Year)") +
  ggtitle("Relationship between First Round Picks in Consecutive Years") 
```

```{r}
ggplot(pickComps, aes(x = diff)) +
  geom_bar() +
  labs(x = "Difference in First Round Slot", y = "Count") +
  ggtitle("Difference between NBA Team's Highest First Round Slot in Consecutive Years") +
   theme_minimal()
```
## Part 2 Analytical Acumen

(A) Prompt: Analyze draft position value and team success/deficiencies compared to expectation.
a. Create a method for valuing each draft slot in the NBA Draft (picks 1 through 60 in most drafts).
b. Conditional on the expected value of the draft positions, which NBA teams have over or
underperformed the most when drafting during this time span. Which College Teams have had the players
outperform expectations the most after entering the NBA?
c. Explain and present your findings with tables and visuals. What additional research areas would you
focus on if given the opportunity to expand this study?


```{r}
# get cumulative win shares for each draft pick
winSharesDF = df %>%
  group_by(overall_pick) %>%
  summarise(TotalWinShares = sum(win_shares, na.rm = T))
# plot cumulative win shares by each draft pick slot
ggplot(winSharesDF, aes(x = overall_pick, y = TotalWinShares)) +
  geom_bar(stat = "identity", fill = "red") +
  labs(x = "Pick Number", y = "Total Win Shares") +
  ggtitle("Total Win Shares by Pick Value")
```

```{r}
# get cumulative vorp for each draft pick
vorpDF = df %>%
  group_by(overall_pick) %>%
  summarise(TotalVORP = sum(value_over_replacement, na.rm = T))
# plot cumulative vorp by each draft pick slot
ggplot(vorpDF, aes(x = overall_pick, y = TotalVORP)) +
  geom_bar(stat = "identity", fill = "red") +
  labs(x = "Pick Number", y = "Total Value Over Replacment") +
  ggtitle("Total VORP by Pick Value")
```

Let's get the expected value for each pick by getting the mean metric for that pick across
all drafts.

```{r}
# Win Share Version, get mean of win shares for each draft pick slot across all players
winSharesMeanDF = df %>%
  group_by(overall_pick) %>%
  summarise(AVGWinShares = mean(win_shares, na.rm = T))
# display expected value from each draft pick
ggplot(winSharesMeanDF, aes(x = overall_pick, y = AVGWinShares)) +
  geom_bar(stat = "identity", fill = "red") +
  labs(x = "Pick Number", y = "AVG Win Shares") +
  ggtitle("Expected Value by Draft Position") +  theme_minimal()
```

```{r}
# VORP Version, get mean VORP for each draft pick slot across all players
vorpMeanDF = df %>%
  group_by(overall_pick) %>%
  summarise(AVGVORP = mean(value_over_replacement, na.rm = T))
# display expected value from each draft pick
ggplot(vorpMeanDF, aes(x = overall_pick, y = AVGVORP)) +
  geom_bar(stat = "identity", fill = "red") +
  labs(x = "Pick Number", y = "AVG Career VORP") +
  ggtitle("Expected Value by Draft Position") +  theme_minimal()
```

```{r}
library(knitr)
library(kableExtra)
# display the expected value in terms of VORP for each pick
kable(vorpMeanDF, "html") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```


```{r}
# add expected value dataframes to original dataframe
df = merge(df, winSharesMeanDF, by = "overall_pick")
df = merge(df, vorpMeanDF, by = "overall_pick")
# define VORP and WS Value, difference between true and expected metrics
df$pickValueWS = df$win_shares - df$AVGWinShares
df$pickValueVORP = df$value_over_replacement - df$AVGVORP
```

```{r}
# display the WS Value for each team over their history of drafting
wsPickValueDF = df %>%
  group_by(team) %>%
  summarise(totalWSValue = sum(pickValueWS, na.rm = T)) %>% 
  arrange(desc(totalWSValue))
kable(wsPickValueDF, "html") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
#ggplot(wsPickValueDF, aes(x = team, y = totalWSValue)) +
#  geom_bar(stat = "identity", fill = "red") +
#  labs(x = "Team", y = "Win Shares Value") +
#  ggtitle("Team's Return on Drafted Expected Value")
```

```{r}
# display the VORP Value for each team over their history of drafting
vorpPickValueDF = df %>%
  group_by(team) %>%
  summarise(totalVORPValue = sum(pickValueVORP, na.rm = T)) %>% arrange(desc(totalVORPValue))
kable(vorpPickValueDF, "html") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```


```{r}
# display the VORP Value for each college over their history of drafting
vorpPickValueDF = df %>%
  group_by(college) %>%
  summarise(totalVORPValue = sum(pickValueVORP, na.rm = T)) %>% arrange(desc(totalVORPValue))
kable(vorpPickValueDF, "html") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"))
```








 